Visualizing Independence Using Extended Association Plots

### 2005-08-25 trunk/vcd/paper/JCGS.Rnw
JCGS paper finished
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@446 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% Packages:
\usepackage{graphicx,color,a4wide,thumbpdf,hyperref}

%% \usepackage{Sweave} is essentially
\RequirePackage[T1]{fontenc}
\RequirePackage{ae,fancyvrb}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontshape=sl}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{}
\DefineVerbatimEnvironment{Scode}{Verbatim}{fontshape=sl}
\newenvironment{Schunk}{}{}

@@

%% Hyperref
\hypersetup{%
  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {true},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

\begin{abstract}
Residual-based shadings for enhancing mosaic and association plots
to visualize independence models for contingency tables are extended
in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
result of an associated significance test is coded by the presence of
color in the visualization. For obtaining (a), a general strategy for
deriving diverging palettes in the perceptually-based HCL space is suggested. As for (b),
cut offs that control the appearance of color are computed in a data-driven
way based on the conditional permutation distribution of maximum-type
test statistics. The shadings are first established for the case of independence
in 2-way tables and then extended to more general independence models
for multi-way tables, including in particular the conditional independence
problem.\\[0.3cm]

@@

Relationships between categorical variables are typically
analyzed based on the underlying contingency tables that can be explored
for (in)dependence. Two standard methods from the statistical tool box
are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
independence and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
bringing it out graphically. Both methods can also be combined such that
a certain mosaic plot visualizes a particular log-linear model by 
controlling splitting order and direction and shading of the tiles
in the mosaic display
\citep{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
In particular, \cite{vcd:Friendly:1994} suggested a shading strategy
based on the residuals (typically, Pearson or deviance residuals) of the associated
log-linear model that elevates the mosaic plot from a display for frequencies
in a contingency table to a visualization technique that encompasses both, observed
frequencies and residuals. This shading allows for judging the quality of a model
fit and spotting dependence patterns that have not been accounted for by the model.
Other techniques for visualizing independence in contingency tables such as
association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
residual-based shading.

In this paper, the shading of \cite{vcd:Friendly:1994} is extended in two directions:
usage of perceptually-based HCL (Hue-Chroma-Luminance) colors and combination of
visualization and significance
testing. Friendly's shading is typically implemented in statistical software using color
spaces whose dimensions are only poorly mapped to the perceptual dimensions of the
human visual system \citep{vcd:Brewer:1999,vcd:Ihaka:2003} which makes it more difficult
to properly read and interpret the corresponding plots. For overcoming this problem,
a general strategy for constructing diverging palettes in the perceptually-based HCL
space \citep{vcd:Ihaka:2003} is derived. To couple the visualization and the independence
model of a contingency table more tightly than in previous approaches, the residual-based
shading is extended such that presence of color in the display is equivalent to significance
of the associated independence test. This is achieved by using data-driven cut offs
for the appearance of color computed from the conditional permutation distribution
\citep{vcd:Ernst:2004,vcd:Pesarin:2001} of maximum-type test statistics. The resulting
residual-based shadings are illustrated both for mosaic and association plots 
using real world data sets.

The remainder of the paper is structured as follows: Section~\ref{sec:2way} gives a
brief introduction to significance tests and visualization techniques for the
independence problem in 2-way contingency tables. Based on this, Section~\ref{sec:shadings}
introduces the extended residual-based shadings using perceptually uniform HCL colors
and combining visualization and significance testing. The results are generalized to
multi-way tables in Section~\ref{sec:3way} with a more detailed discussion of adapting
them to conditional independence. Section~\ref{sec:conclusions} summarizes the paper
and gives some concluding remarks.

@@

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced first in statistics
textbooks if they address the independence problem in 2-way tables
\citep[see, e.g.,][]{vcd:Agresti:2002}.

However, the sum of squares is not the only plausible way of capturing deviations
from $0$ in the residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
suitable for identifying the cells responsible for the dependence (if any) is
the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
for the dependence can easily be identified.

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic from Equation~(\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep[e.g.,][]{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from Equation~(\ref{eq:H0}), using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

@@

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from Equation~(\ref{eq:ChiSquare})
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistic from Equation~(\ref{eq:MaxAbs}) is $M = \Sexpr{stat(art_max)}$
with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
from a sample of size $5000$ from the permutation distribution under independence
generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the \cite{vcd:Patefield:1981} algorithm and computing the maximum statistic for 
each of these tables.

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS (Hue-Luminance-Saturation) color space.
The implementation in the \proglang{R} system for statistical computing and graphics
\citep{vcd:R:2005} employs colors from HSV (Hue-Saturation-Value) space. 
Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. As both spaces can generate the
same colors, we only discuss HSV space in the following.

@@

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ from Equation~(\ref{eq:MaxAbs}) which always employs the critical values $c_\alpha$
$2$ and $4$. The problem is that it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$, respectively, and exactly the cells which
violate the independence hypothesis are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

@@

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[htbp]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the arthritis data with Friendly shading (left) and maximum shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

\section{Extension to Conditional Independence} \label{sec:3way}

@@

As these independence models for multi-way tables also provide the structure
required for the residual-based HCL shadings derived in Section~\ref{sec:shadings},
both the maximum shading and the sum-of-squares shading can straightforwardly be applied.
However, such independence models often additionally provide further structure that allows
decomposition of the overall model into smaller independence problems which can be exploited
both for choosing appropriate data-driven cut-offs in the shading and for selecting 
a suitable layout of mosaic or association plots. Specific strategies for the conditional
independence problem are derived in the following.

@@

Alternatively, a maximum statistic similar to that from Equation~(\ref{eq:MaxAbs})
can be constructed
\begin{equation} \label{eq:doublemax}
\max_{i, j, k} |r_{ijk}| \quad = \quad \max_k M_k.
\end{equation}
As in the 2-way case, this allows for identification of the cells that are responsible
for the deviation from conditional independence (if any). If it is not so much of interest
in which \emph{cell} but only in which \emph{stratum} the deviation occurs, then
it would be natural to use 
\begin{equation} \label{eq:maxChisq}
\max_k \sum_{i,j} r_{ijk}^2 \quad = \quad \max_k X_k^2.
\end{equation}
Given a critical value for this statistic, all strata $k$ whose associated sum-of-squares
statistics $X_k^2$ exceed the critical value are in conflict with the hypothesis of 
conditional independence.

All these statistics are of type
\begin{equation} \label{eq:coindep_test}
\lambda_\mathrm{agg} (\lambda_\mathrm{indep}([r_{ijk}])),
\end{equation}
where $\lambda_\mathrm{indep}$ is a functional for assessing independence in stratum $k$
and $\lambda_\mathrm{agg}$ is a functional for aggregating over the $k = 1, \dots, K$ strata.
If the maximum is used for the latter, then identification of the strata responsible for the
dependence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
the corresponding cells can also be identified. Hence, the double maximum
statistic from Equation~(\ref{eq:doublemax}) is the only functional allowing for detection of
both the strata and the cells violating the conditional independence hypothesis.

However, the main purpose of the formulation of the different test statistics is not so much 
inference but their applicability to diagnostic plots via residual-based shadings. As already
discussed, it is possible for all aggregation functionals to simply use either
the significant or the non-significant shading for all cells in the contingency table---this
strategy would have to be used for the sum-of-squares statistic from Equation~(\ref{eq:sumChisq}). If
$\lambda_\mathrm{agg}$ is the maximum as in Equation~(\ref{eq:maxChisq}),
then the significant palette would only be used
in those strata in conflict with the hypothesis of conditional independence whereas the
non-significant palette would be used for the remaining strata. Finally, if both $\lambda_\mathrm{agg}$
and $\lambda_\mathrm{indep}$ are the maximum, then the same strategy as in Section~\ref{sec:shadings}
can be pursued, i.e., only the significant palette is used but with data-driven cut offs
derived from the distribution of the double maximum statistic from Equation~(\ref{eq:doublemax}).

@@

For illustration, the famous admissions data of the University of
California, Berkley, (UCB) from \cite{vcd:Bickel+Hammel+O'Connell:1975}
are employed. Testing the conditional independence of admission and gender
given the department leads to the well-known result that there is no 
gender discrimination in five of six departments, but that women are more
likely to be admitted than men in the first department. This departure
from conditional independence is picked up by all three statistics formulated
above: the sum-of-squares statistic from Equation~(\ref{eq:sumChisq}) is
$\sum_k X_k^2 = \Sexpr{stat(ucb_sumChisq)}$ ($p = \Sexpr{pval(ucb_sumChisq)}$),
the maximum sum-of-squares statistic from Equation~(\ref{eq:maxChisq}) is
$\max_k X_k^2 = \Sexpr{stat(ucb_maxChisq)}$ ($p = \Sexpr{pval(ucb_maxChisq)}$),
and the double maximum statistic from Equation~(\ref{eq:doublemax}) is
$\max_k M_k = \Sexpr{stat(ucb_doublemax)}$ ($p = \Sexpr{pval(ucb_doublemax)}$).
The $p$ values are again computed by drawing $5000$ samples from the corresponding
permutation distribution. The latter two test statistics both indicate the 
first department as the source of departure and the maximum statistic additionally
identifies the deviations in the female applicants group. This result
is also visualized by the conditional association plot in a trellis layout based
on the double maximum shading in Figure~\ref{fig:ucb}. This can be
interpreted as a diagnostic residual plot for the associated log-linear model
because it shows that there are significant departures from the model in department
A but not in departments B to F where all residuals are very small.

@@

The results in this paper were obtained using
\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")}
\citep[\mbox{\url{http://www.R-project.org/}}]{vcd:R:2005} and the packages
\pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2005},
{\pkg{MASS}~\Sexpr{packageDescription("MASS")$Version}} \citep{vcd:Venables+Ripley:2002},
{\pkg{grid}~{\Sexpr{packageDescription("grid")$Version}}} \citep{vcd:Murrell:2002} and
{\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.

>>>>

%% Packages:
\usepackage{graphicx,color,a4wide,thumbpdf,hyperref}
%% need no \usepackage{Sweave}

@@

%% Hyperref
\hypersetup{%
  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

@@

\begin{abstract}
Residual-based shadings for enhancing mosaic and association plots
to visualize independence models for contingency tables are extended
in two directions: (a)~perceptually uniform HCL colors are used and (b)~the
result of an associated significance test is coded by the appearance of
color in the visualization. For obtaining (a), a general strategy for
deriving diverging palettes in the perceptually-based HCL space is suggested. As for (b),
cut offs that control the appearance of color are computed in a data-driven
way based on the conditional permutation distribution of maximum-type
test statistics. The shadings are first established for the case of independence
in 2-way tables and then extended to more general independence models
for multi-way tables, including in particular conditional independence
problems.\\[0.3cm]

@@

Relationships between categorical variables are typically
analyzed based on the underlying contingency tables that can be explored
for (in)dependence. Two standard methods from the statistical tool box
are log-linear models \citep[see, e.g.,][]{vcd:Agresti:2002} for modeling
(in)dependence structures and mosaic plots \citep{vcd:Hartigan+Kleiner:1981} for 
bringing them out graphically. Both methods can also be combined such that
a certain mosaic plot visualizes a particular log-linear model by 
controlling splitting order and direction and shading of the tiles
in the mosaic display
\citep{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
In particular, \cite{vcd:Friendly:1994} suggested a shading strategy
based on the residuals (typically, Pearson or deviance residuals) of the associated
log-linear model that elevates the mosaic plot from a display for frequencies
in a contingency table to a visualization technique that encompasses both, observed
frequencies and residuals. This shading allows for judging the quality of a model
fit and spotting dependence patterns that have not been accounted for by the model.
Other techniques for visualizing independence in contingency tables such as
association plots \citep{vcd:Cohen:1980} can also be enhanced by using this
residual-based shading.

In this paper, the shading of \cite{vcd:Friendly:1994} is extended in two directions:
usage of perceptually-based HCL (Hue-Chroma-Luminance) colors and combination of
visualization and significance
testing. Friendly's shading is typically implemented in statistical software using color
spaces whose dimensions are only poorly mapped to the perceptual dimensions of the
human visual system \citep{vcd:Brewer:1999,vcd:Ihaka:2003} which makes it more difficult
to properly read and interpret the corresponding plots. For overcoming this problem,
a general strategy for constructing diverging palettes in the perceptually-based HCL
space \citep{vcd:Ihaka:2003} is derived. To couple the visualization and the independence
model of a contingency table more tightly than in previous approaches, the residual-based
shading is extended such that appearance of color in the display is equivalent to significance
of the associated independence test. This is achieved by using data-driven cut offs
for the appearance of color computed from the conditional permutation distribution
\citep{vcd:Ernst:2004,vcd:Pesarin:2001} of maximum-type test statistics. The resulting
residual-based shadings are illustrated both for mosaic and association plots 
using real world data sets.

The remainder of the paper is structured as follows: Section~\ref{sec:2way} gives a
brief introduction to significance tests and visualization techniques for the
independence problem in 2-way contingency tables. Based on this, Section~\ref{sec:shadings}
introduces the extended residual-based shadings using perceptually uniform HCL colors
and combining visualization and significance testing. The results are generalized to
multi-way tables in Section~\ref{sec:3way} with a more detailed discussion of adapting
them to conditional independence problems. Section~\ref{sec:conclusions} summarizes the paper
and gives some concluding remarks.

@@

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n_{++}$. The probably best known and most used
measure of discrepancy between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have an unconditional limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test which is typically introduced first in statistics
textbooks when addressing the independence problem in 2-way tables
\citep[see, e.g.,][]{vcd:Agresti:2002}.

However, the sum of squares is not the only plausible way of capturing deviations
from zero in the residuals.
There are many other conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda([r_{ij}])$: one which is particularly
suitable for identifying the cells responsible for the `dependence' (i.e., significant
departure from independence), if any, is the maximum of the absolute values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the null hypothesis of indendence at significance level
$\alpha$ \cite[][ch.~7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells responsible
for the dependence can easily be identified.

Furthermore, an important reason for using the unconditional limiting distribution
for the $X^2$ statistic from Equation~\ref{eq:ChiSquare} was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \citep[e.g.,][]{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from Equation~\ref{eq:H0}, using a permutation test is particularly
intuitive due to the permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

@@

For the arthritis data from Table~\ref{tab:art}, both tests indicate a clearly significant
dependence of improvement on treatment: the sum-of-squares statistic from Equation~\ref{eq:ChiSquare}
is $X^2 = \Sexpr{stat(art_chisq)}$ with a $p$ value of $p = \Sexpr{pval(art_chisq)}$,
and the maximum statistic from Equation~\ref{eq:MaxAbs} is $M = \Sexpr{stat(art_max)}$
with $p = \Sexpr{pval(art_max)}$. Both $p$ values have been computed
from a sample of size $5000$ from the permutation distribution under independence
generated via sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the \cite{vcd:Patefield:1981} algorithm and computing the maximum statistic for 
each of these tables.

@@

The way the (light) blue and red colors are chosen differs somewhat between
various implementations of Friendly mosaic plots: In his original 
\proglang{SAS} implementation \citep[see][]{vcd:Friendly:2000}, Michael Friendly
uses colors from a palette based on HLS (Hue-Luminance-Saturation) color space.
The implementation in the standard packages of the
\proglang{R} system for statistical computing and graphics
\citep{vcd:R:2005} employs colors from HSV (Hue-Saturation-Value) space. 
Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\citep{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \citep{vcd:Moretti+Lyons:2002} making the generation
of the Friendly shading very simple. As both spaces can generate the
same colors, we only discuss HSV space in the following.

@@

The Friendly shading can be interpreted to be a visualization of the maximum
statistic $M$ from Equation~\ref{eq:MaxAbs} which always employs the critical values $c_\alpha$
$2$ and $4$. However, it is not clear to which
significance levels $\alpha$ these critical values correspond because the distribution
of $M$ depends on the underlying contingency table. The natural solution to this
problem is to compute the critical values from the distribution of $M$ in a data-driven
way (i.e., for the table visualized) and use these instead of the hard-coded values
$2$ and $4$. In the right panel of Figure~\ref{fig:shadings} this is done for
the arthritis data by employing the critical values
$\Sexpr{round(as.vector(art_max$qdist(0.9)), digits = 2)}$ at level $\alpha = 0.1$ and
$\Sexpr{round(as.vector(art_max$qdist(0.99)), digits = 2)}$ at level $\alpha = 0.01$
(and using the diverging HCL palette) derived from the permutation distribution of $M$ for 
the arthritis data as described in Section~\ref{sec:tests}. By using these cut offs, the presence of color
in the plot is equivalent to significance (of the maximum statistic $M$) at 
level $\alpha = 0.1$ and $\alpha = 0.01$, respectively, and exactly the cells which
violate the independence hypothesis are highlighted. For the arthritis data, these are in particular the 
cells in the last column that signal that there are significantly more
treated patients and fewer placebo patients with marked improvement than would be expected
under independence.

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<shadings,fig=TRUE,echo=FALSE,height=6,width=11,results=hide>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
pushViewport(viewport(layout.pos.col=1, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = shading_Friendly, margins = c(2.5, 4, 2.5, 3), gp_args = list(eps = NULL, lty = 1))
popViewport()
pushViewport(viewport(layout.pos.col=2, layout.pos.row=1))
mosaic(art, newpage = FALSE, gp = art_gp, margins = c(2.5, 3, 2.5, 4))
popViewport(2)
@
\caption{Mosaic plot for the arthritis data with Friendly shading (left) and maximum shading (right).}
\label{fig:shadings}
\end{center}
\end{figure}

@@

\section{Extensions} \label{sec:3way}

@@

As these independence models for multi-way tables also provide the structure
required for the residual-based HCL shadings derived in Section~\ref{sec:shadings},
both the maximum shading and the sum-of-squares shading can straightforwardly be applied.
However, such independence models often additionally provide further structure that allows
decomposition of the overall model into smaller independence problems which can be exploited
both for choosing appropriate data-driven cut-offs in the shading and for selecting 
a suitable layout of mosaic or association plots. Specific strategies for the conditional
independence problem in 3-way tables are derived in the following.

@@

Alternatively, a maximum statistic similar to that from Equation~\ref{eq:MaxAbs}
can be constructed
\begin{equation} \label{eq:doublemax}
\max_{i, j, k} |r_{ijk}| \quad = \quad \max_k M_k.
\end{equation}
As in the 2-way case, this allows for identification of the cells that are responsible
for the deviation from conditional independence (if any). If it is not so much of interest
in which \emph{cell} but only in which \emph{stratum} the deviation occurs, then
it would be natural to use 
\begin{equation} \label{eq:maxChisq}
\max_k \sum_{i,j} r_{ijk}^2 \quad = \quad \max_k X_k^2.
\end{equation}
Given a critical value for this statistic, all strata $k$ whose associated sum-of-squares
statistics $X_k^2$ exceed the critical value are in conflict with the hypothesis of 
conditional independence.

All these statistics are of type
\begin{equation} \label{eq:coindep_test}
\lambda_\mathrm{agg} (\lambda_\mathrm{indep}([r_{ijk}])),
\end{equation}
where $\lambda_\mathrm{indep}$ is a functional for assessing independence in stratum $k$
and $\lambda_\mathrm{agg}$ is a functional for aggregating over the $k = 1, \dots, K$ strata.
If the maximum is used for the latter, then identification of the strata responsible for the
dependence is possible. If additionally $\lambda_\mathrm{indep}$ is the maximum,
the corresponding cells can also be identified. Hence, the double maximum
statistic from Equation~\ref{eq:doublemax} is the only functional allowing for detection of
both the strata and the cells violating the conditional independence hypothesis.

However, the main purpose of the formulation of the different test statistics is not so much 
inference but their applicability to diagnostic plots via residual-based shadings. As already
discussed, it is possible for all aggregation functionals to simply use either
the significant or the non-significant shading for all cells in the contingency table---this
strategy would have to be used for the sum-of-squares statistic from Equation~\ref{eq:sumChisq}. If
$\lambda_\mathrm{agg}$ is the maximum as in Equation~\ref{eq:maxChisq},
then the significant palette would only be used
in those strata in conflict with the hypothesis of conditional independence whereas the
non-significant palette would be used for the remaining strata. Finally, if both $\lambda_\mathrm{agg}$
and $\lambda_\mathrm{indep}$ are the maximum, then the same strategy as in Section~\ref{sec:shadings}
can be pursued, i.e., only the significant palette is used but with data-driven cut offs
derived from the distribution of the double maximum statistic from Equation~\ref{eq:doublemax}.

@@

For illustration, the famous admissions data of the University of
California, Berkley, (UCB) from \cite{vcd:Bickel+Hammel+O'Connell:1975}
are employed. Testing the conditional independence of admission and gender
given the department leads to the well-known result that there is no 
gender discrimination in five of six departments, but that women are more
likely to be admitted than men in the first department. This departure
from conditional independence is picked up by all three statistics formulated
above: the sum-of-squares statistic from Equation~\ref{eq:sumChisq} is
$\sum_k X_k^2 = \Sexpr{stat(ucb_sumChisq)}$ ($p = \Sexpr{pval(ucb_sumChisq)}$),
the maximum sum-of-squares statistic from Equation~\ref{eq:maxChisq} is
$\max_k X_k^2 = \Sexpr{stat(ucb_maxChisq)}$ ($p = \Sexpr{pval(ucb_maxChisq)}$),
and the double maximum statistic from Equation~\ref{eq:doublemax} is
$\max_k M_k = \Sexpr{stat(ucb_doublemax)}$ ($p = \Sexpr{pval(ucb_doublemax)}$).
The $p$ values are again computed by drawing $5000$ samples from the corresponding
permutation distribution. The latter two test statistics both indicate the 
first department as the source of departure and the maximum statistic additionally
identifies the deviations in the female applicants group. This result
is also visualized by the conditional association plot in a trellis layout based
on the double maximum shading in Figure~\ref{fig:ucb}. This can be
interpreted as a diagnostic residual plot for the associated log-linear model
because it shows that there are significant departures from the model in department
A but not in departments B to F where all residuals are very small.

@@

The results in this paper were obtained using
\proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")}
\citep[\mbox{\url{http://www.R-project.org/}}]{vcd:R:2005} and the packages
\pkg{vcd}~\Sexpr{packageDescription("vcd")$Version} \citep{vcd:Meyer+Zeileis+Hornik:2005},
{\pkg{MASS}~\Sexpr{packageDescription("MASS")$Version}} \citep{vcd:Venables+Ripley:2002},
{\pkg{grid}~{\Sexpr{packageDescription("grid")$Version}}} \citep[see][]{vcd:Murrell:2002} and
{\pkg{colorspace}~{\Sexpr{packageDescription("colorspace")$Version}}} \citep{vcd:Ihaka:2004}.
